---
title: "Benchmark Results"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## 0. Setup 
```{r}
library(tidyverse)
library(cowplot)

# Import results
results_df <- read_csv("data/results/combined_results.csv")
```

## 1. Test 1 - Varying Fit Size for Fixed Prediction Size
```{r}
results_test1 <- results_df %>%
    filter(test == 1) %>%
    filter(elements %in% c(2, 3)) %>%
    filter(n_train > 2^9)

func_log2 <- function(x) paste0("2^", log2(as.numeric(x)))

results_test1 %>%
    ggplot(aes(
        x = n_train,
        y = time_pred,
        color = model
    )) +
    geom_point(size = 3) +
    facet_wrap(~elements) +
    scale_x_continuous(
        breaks = unique(results_test1$n_train),
        trans = "log2",
        labels = func_log2
    ) +
    labs(
        x = "Number of Training Samples",
        y = "Fit Time (sec)",
        color = "Model",
        title = "Test 1: Increasing Fitting Size for Fixed Prediction Size"
    ) +
    theme_cowplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
ggsave("test1_plot.pdf", width = 12, height = 7, units = "in")
```

## 2. Test 2 - Increasing Fit Size
```{r}
results_test2 <- results_df %>%
    filter(test == 2)

results_test2 %>%
    ggplot(aes(
        x = as.factor(n_train),
        y = as.factor(n_covs),
        fill = time_fit
    )) +
    geom_tile() +
    scale_x_discrete(
        labels = func_log2
    ) +
    facet_grid(
        elements ~ model,
        labeller = labeller(
            elements = as_labeller(c(
                "2" = "2 Elements",
                "3" = "3 Elements",
                "4" = "4 Elements",
                "5" = "5 Elements"
            )), model = label_value
        )
    ) +
    scale_fill_viridis_c(option = "plasma") +
    labs(
        x = "Number of Training Samples",
        y = "Number of Covariates",
        fill = "Fit Time (sec)",
        title = "Test 2: Increasing Fit Size"
    ) +
    theme_cowplot()
```

## 3. Test 3 - Prediction Parallelisation
### 3.1 Weak Scaling 
```{r}
results_test31 <- results_df %>%
    filter(test == 3 & time_pred != -1) %>%
    filter((n_test / cpus) == 16) %>%
    group_by(model, elements, n_covs) %>%
    mutate(t1 = mean(time_pred[cpus == 1], na.rm = TRUE)) %>%
    ungroup() %>%
    group_by(model, elements, n_covs, cpus) %>%
    mutate(tn = mean(time_pred), na.rm = TRUE) %>%
    ungroup() %>%
    mutate(efficiency = t1 / tn) %>%
    filter(elements == 2)

results_test31 %>%
    ggplot(aes(
        x = cpus,
        y = efficiency,
        color = model
    )) +
    facet_wrap(~elements) +
    geom_line(size = 1) +
    scale_x_continuous(
        breaks = unique(results_test31$cpus),
        trans = "log2", labels = func_log2
    ) +
    geom_hline(
        aes(yintercept = 1, color = "ideal"),
        linetype = "dashed", size = 1, color = "gray"
    ) +
    labs(
        x = "Number of CPUs",
        y = "Efficiency",
        color = "Model",
        title = "Test 3.1: Weak Scaling"
    ) +
    theme_cowplot()
ggsave("test31_plot.pdf", width = 12, height = 7, units = "in")
```

### 3.2 Strong Scaling
```{r}
results_test32 <- results_df %>%
    filter(test == 3 & time_pred != -1) %>%
    group_by(model, elements, n_covs) %>%
    mutate(t1 = mean(time_pred[cpus == 1], na.rm = TRUE)) %>%
    ungroup() %>%
    group_by(model, elements, n_covs, cpus) %>%
    mutate(tn = mean(time_pred), na.rm = TRUE) %>%
    ungroup() %>%
    mutate(efficiency = t1 / tn) %>%
    filter(!is.na(efficiency))

p_values <- seq(2^0, 2^6, by = 1) # Range of CPUs
amdahl <- function(p, f) {
    1 / ((1 - f) + (f / p))
}
amdahl_curves <- rbind(
    data.frame(p = p_values, speedup = amdahl(p_values, 0.96), f = "f = 0.96"),
    data.frame(p = p_values, speedup = amdahl(p_values, 0.90), f = "f = 0.90"),
    data.frame(p = p_values, speedup = amdahl(p_values, 0.45), f = "f = 0.55"),
    data.frame(p = p_values, speedup = amdahl(p_values, 1), f = "f = 1.0")
)

results_test32 %>%
    ggplot(aes(
        x = cpus,
        y = efficiency,
        color = model
    )) +
    facet_wrap(~elements) +
    geom_line(size = 1) +
    geom_line(
        data = amdahl_curves,
        aes(x = p, y = speedup, linetype = f),
        color = "gray"
    ) +
    scale_linetype_manual(values = c("dashed", "dashed", "dashed", "dashed")) +
    scale_x_continuous(
        breaks = unique(results_test32$cpus),
        trans = "log2"
    ) +
    labs(
        x = "Number of CPUs",
        y = "Efficiency",
        color = "Model",
        title = "Test 3.2: Strong Scaling"
    ) +
    theme_cowplot()
ggsave("test32_plot.pdf", width = 12, height = 7, units = "in")
```

# 4. Test 4 - Coarsening 
```{r}
results_test4 <- results_df %>%
    filter(test == 4)

results_test4 %>%
    ggplot(aes(
        x = times,
        y = time_pred,
        color = as.factor(n_test)
    )) +
    geom_point(size = 3) +
    facet_wrap(~model, scales = "free_y", nrow = 3) +
    scale_x_continuous(
        breaks = unique(results_test4$times),
        trans = "log2",
        labels = func_log2
    ) +
    labs(
        x = "Number of Distinct Event Times",
        y = "Fit Time (sec)",
        color = "N(Pred)",
        title = "Test 4: Increasing Number of Distinct Event Times"
    ) +
    theme_cowplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))


ggsave("test4_plot.pdf", width = 12, height = 7, units = "in")
```